Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-11-20 rnni_convexity.tex
up to thm caterpillar convexity

In this section we restrict our attention to the set of caterpillar trees, which we will represent throughout this section in the simplified list representation as introduced in Section~\ref{section:alg_csort}.
This type of tree is of particular interest in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
This in combination with the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{Dasgupta2000-xa} suggests that understanding the geometry of paths between caterpillar trees helps investigating the complexity of computing $\rnni$ distances.
In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
We discovered this by computing caterpillar paths with $\csort$ and comparing them with shortest path in $\rnni$, computed by the exact algorithm discussed in Section~\ref{section:alg_RNNI_graph}.
In the following we demonstrate why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.

Therefore, we compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
In $\nni$, building a cherry and moving it may give a path that is shorter than a path where the taxa move separately in as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
There is only one move needed to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon in $\nni$.
But in $\rnni$ space there are additional rank moves needed to change the rank of the internal node of the cherry before an $\nni$ move can resolve it.
This is due to the fact that $\nni$ moves are only allowed on edges of length one in $\rnni$.
Hence a path where taxa move separately and no second cherry is built, is shortest in $\rnni$.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{NNI_vs_RNNI}
\vspace{12pt}
\caption{Paths between caterpillar trees $T$ and $R$: The solid paths are paths in $\rnni$, the dashed one is a path that is only possible in the plain $\nni$ graph.
The upper path, which is the $\rnni$ path that is the natural extension of the shortest $\nni$ path, is longer than a shortest $\rnni$ path as the one at the bottom.}
\label{fig:NNI_vs_RNNI}
\end{figure}

The above observation suggests that there is always a caterpillar path that is a shortest path between any two caterpillar trees in $\rnni$.
Indeed, we are going to prove Theorem~\ref{thm:caterpillar_convex} within this section, which states that the set of caterpillar trees is convex in $\rnni$.
Note that this is not true for the $\nni$ space, which is clear from the example in Figure~\ref{fig:NNI_vs_RNNI}.

Lemma~\ref{lemma:caterpillar_dist=diameter} regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.
Notice that $\frac{(n-1)(n-2)}{2}$ is the upper bound for the diameter of $\rnni$ by Lemma~\ref{lemma:diameter_bound}.

Within this chapter we will again use the list representation for caterpillar trees, which is a simplified version of the cluster representation of trees, as explained in Section~\ref{section:alg_csort}.

@@

\begin{proof}
Let $T$ and $R$ be caterpillar trees.
Let us first prove that if $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
This easily follows from algorithm $\csort$:
In each loop $i=1, \ldots, n-2$ of Algorithm~\ref{alg:csort} there are at most $n-1-i$ exchanges of taxa.
This is due to the fact that after loop number $i$ the last $i$ elements in the list representation of the current tree on $p$ are sorted correctly as in the destination tree.
Hence there are no more exchanges of pairs of taxa needed within the last $i-1$ of the sequences in loop $i$, and so only $n-1-i$ swaps of elements might appear within the first $n-i$ elements of the list.
It follows that the maximum length of a path computed by $\csort$ is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
Since there cannot be a caterpillar path shorter than $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.

@@

Induction basis: $n=3$\\
There exist three trees on three taxa: $[1,2,3], [1,3,2]$ and $[2,3,1]$.
They all have distance one from another and are caterpillar trees.
So it is $d(T,R) = d_c(T,R) = 1$ for all pairs of trees $T,R$ on $n=3$ taxa.

Induction step: $n \to n+1$\\
Induction hypothesis: For caterpillar trees $T, R$ on $n$ taxa with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$ it is $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
Without loss of generality we can assume that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$.
Let $R$ be a caterpillar tree with distance $\frac{n(n-1)}{2}$ to $T$.
Then each of the $\frac{n(n-1)}{2}$ pairs of taxa that is considered when running $\csort$ to compute a path $p$ from $T$ to $R$ exchanges.
In particular, taxon $n+1$ exchanges with a taxon in each of the $n+1$ loops in $\csort$.
We can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$.
Let us now consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
$\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
Hence, the caterpillar distance is $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.

We are now using this observation to prove that the distance between $T$ and $R$ equals $\frac{n(n-1)}{2}$.
It is obvious that it gives an upper bound for the distance as it is $\Delta(\rnni_{n+1}) \leq \frac{n(n-1)}{2}$ for the $\rnni$ graph $\rnni_{n+1}$ on $n+1$ taxa by Lemma~\ref{lemma:diameter_bound}.
So we now need to prove that $\frac{n(n-1)}{2}$ is a lower bound for the distance between $T$ and $R$ as well.
According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T{\big|}_n$ and $R{\big|}_n$ is at least $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.
We also know $|\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| = n-1$.
Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\frac{n(n-1)}{2}$.
By the above, it follows that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
Since this is a contradiction to $d(T{\big|}_n, R{\big|}_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.
\end{proof}

>>>>

In this section we restrict our attention to the set of caterpillar trees.
Throughout this section the simplified list representation for these trees as introduced in Section~\ref{section:alg_csort} will be used.
This type of tree is of particular interest in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
Following this observation and the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{Dasgupta2000-xa}, understanding the geometry of paths between caterpillar trees seems essential for investigating the complexity of computing $\rnni$ distances.
In the following we demonstrate why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.

Let us compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
It can happen in $\nni$ that building a cherry and moving it may give a path that is shorter than a path where the taxa move separately in as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
There is only one move needed to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon in $\nni$.
In $\rnni$ space however, there are additional rank moves needed to change the rank of the internal node of the cherry before an $\nni$ move can resolve it.
This is due to the fact that $\nni$ moves are only allowed on edges of length one in $\rnni$.
As a consequence a path where taxa move separately and no new cherry is built is shortest in $\rnni$.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{NNI_vs_RNNI}
\vspace{12pt}
\caption{Paths between caterpillar trees $T$ and $R$: The solid paths are paths in $\rnni$, the dashed one is a path that is only possible in the plain $\nni$ graph.
The upper path is a natural extension of the shortest $\nni$ path in $\rnni$.
It is longer than a shortest $\rnni$ path as the one at the bottom which consists of caterpillar trees only.}
\label{fig:NNI_vs_RNNI}
\end{figure}

The above observation suggests that there is a caterpillar path that is a shortest path between any two caterpillar trees in $\rnni$.
Indeed, we are going to prove within this section that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}).
Note that this is not true for the $\nni$ space, which is clear from the example in Figure~\ref{fig:NNI_vs_RNNI}.

The following lemma regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.
Notice that $\frac{(n-1)(n-2)}{2}$ is the upper bound for the diameter of $\rnni$ by Lemma~\ref{lemma:diameter_bound}.

@@

\begin{proof}
Let $T$ and $R$ be caterpillar trees.
Let us first prove that if $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
This easily follows from algorithm $\csort$:
In each loop $i=1, \ldots, n-2$ of $\csort$ (Algorithm~\ref{alg:csort}) there are at most $n-1-i$ pairs of taxa that swap positions.
This is due to the fact that after loop number $i$ the last $i$ elements in the list representation of the current tree on $p$ are in the same place as in the destination tree.
As a result there are no further exchanges of pairs of taxa within the last $i-1$ taxa of the tree in loop $i$, which means that at most $n-1-i$ pairs of taxa exchange then.
It follows that the maximum length of a path computed by $\csort$ is $\sum\limits_{i=1}^{n-2} n-1-i = \frac{(n-1)(n-2)}{2}$.
Since there cannot be a caterpillar path shorter than $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.

@@

Induction basis: $n=3$\\
There are three trees on three taxa.
They all have distance one from another and are caterpillar trees.
So it is $d(T,R) = d_c(T,R) = 1$ for all pairs of trees $T,R$ on $n=3$ taxa.

Induction step: $n \to n+1$\\
Induction hypothesis: For caterpillar trees $T, R$ on $n$ taxa with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$ it is $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
Without loss of generality we can assume that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$.
Let $R$ be a caterpillar tree with distance $\frac{n(n-1)}{2}$ to $T$.
It follows that in each loop of $\csort$ the maximum number of taxon swaps is reached which implies that every taxon will swap with taxon $n+1$ at some point.
We can conclude that this taxon $n+1$ is part of the cherry of $R$.
Let us now consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
$\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
Hence, the caterpillar distance is $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.

We are now using this observation to prove that the distance between $T$ and $R$ equals $\frac{n(n-1)}{2}$.
This obviously gives an upper bound for the distance as it is $\Delta(\rnni_{n+1}) \leq \frac{n(n-1)}{2}$ for the $\rnni$ graph on $n+1$ taxa by Lemma~\ref{lemma:diameter_bound}.
It is left to prove that $\frac{n(n-1)}{2}$ is a lower bound for the distance between $T$ and $R$ as well.
According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T{\big|}_n$ and $R{\big|}_n$ is at least $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.
We also know $|\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| = n-1$.
Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\frac{n(n-1)}{2}$.
By the above, it follows that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
Since this is a contradiction to $d(T{\big|}_n, R{\big|}_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.
\end{proof}
